---
title: "Efficient list melting and unnesting with rrapply"
author: "Joris Chau"
date: 'Latest date: `r Sys.Date()`'
output:
  html_document:
    toc: no
    df_print: paged
  rmarkdown::html_vignette:
    toc: no
vignette: >
  %\VignetteIndexEntry{Efficient list melting and unnesting with rrapply}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", out.width = "100%")
options(width = 150)
```

# Introduction

The previous article `vignette("1-when-to-use-rrapply")` describes the principal use of the `rrapply()` function as a revised and extended version of base `rapply()` in the context of nested list recursion. For quick data exploration of a nested list it can make sense to keep the list in its original nested format to reduce the number of processing steps and minimize code complexity. As part of a more elaborate data analysis, if there is no specific reason to keep the nested data structure, it is often more practical to transform the nested list into a more convenient rectangular format and work with the unnested object (e.g. a data.frame) instead. In this follow-up article, we review the available (`how`) options in `rrapply()` to unnest or melt nested lists into a rectangular format in more detail and highlight the similarities and differences with respect to several common alternatives in R.

# Nested list to data.frame

## Melt to long data.frame

The option `how = "melt"` in `rrapply()` unnests a nested list to a *long* or melted data.frame similar in format to the retired `reshape2::melt()` function applied to a nested list. The rows of the melted data.frame contain the individual node paths of the elements in the nested list after pruning (based on the `condition` and/or `classes` arguments). The `"value"` column is a vector- or list-column containing the values of the leaf elements identical to the object returned by `how = "flatten"`. 

To demonstrate, we use the `renewable_energy_by_country` dataset included in the `rrapply`-package, a nested list containing the renewable energy shares per country (% of total energy consumption) in 2016[^1]. The 249 countries and areas are structured based on their geographical locations according to the [United Nations M49 standard](https://unstats.un.org/unsd/methodology/m49/). The numeric values listed for each country are percentages, if no data is available the value of the country is `NA`. 

```{r}
library(rrapply)

## melt all data to long data.frame
rrapply(
  renewable_energy_by_country, 
  how = "melt"
) |>
  head(n = 10)
```

```{r}
## drop logical NA's and melt to data.frame
rrapply(
  renewable_energy_by_country,
  classes = "numeric",
  how = "melt"
) |>
  head(n = 10)
```


```{r}
## apply condition and melt to data.frame
rrapply(
  renewable_energy_by_country,
  condition = \(x, .xparents) "Western Europe" %in% .xparents,
  how = "melt"
) |>
  head(n = 10)
```

As shown in the above examples, in comparison to `reshape2::melt()`, `rrapply()` allows to filter or transform list elements before melting the nested list through the `f`, `classes`  and `condition` arguments[^2]. More importantly, `rrapply()` is optimized specifically for handling nested lists, whereas `reshape2::melt()` was aimed primarily at melting data.frames before being superseded by `tidyr::gather()` and more recently `tidyr::pivot_longer()`. For this reason, `reshape2::melt()` can be quite slow when applied to large nested lists:

```{r}
## melt to long data.frame (reshape2)
reshape2::melt(renewable_energy_by_country) |>
  head(10)

## computation times
bench::mark(
  rrapply(renewable_energy_by_country),
  reshape2::melt(renewable_energy_by_country),
  check = FALSE
)
```

For a medium-sized list as used in this example, the computation time of `reshape2::melt()` is not a bottleneck for practical usage. However, the computational effort quickly increases when melting larger or more deeply nested lists:

```{r}
## helper function to generate large nested list
new_list <- function(n, d) {
  v <- vector(mode = "list", length = n)
  rrapply(
    object = v,
    classes = c("list", "NULL"),
    condition = \(x, .xpos) length(.xpos) <= d,
    f = \(x, .xpos) if(length(.xpos) < d) v else runif(1),
    how = "recurse"
  )
}

## generate large shallow list (10^6 elements)
shallow_list <- new_list(n = 100, d = 3)
str(shallow_list, list.len = 2)

## benchmark timing with rrapply
system.time(shallow_melt <- rrapply(shallow_list, how = "melt")) 
head(shallow_melt)

## benchmark timing with reshape2::melt
system.time(shallow_melt_reshape2 <- reshape2::melt(shallow_list))
head(shallow_melt_reshape2)
```

```{r}
## generate large deeply nested list (2^18 elements)
deep_list <- new_list(n = 2, d = 18)

## benchmark timing with rrapply
system.time(deep_melt <- rrapply(deep_list, how = "melt")) 
head(deep_melt)

## benchmark timing with reshape2::melt
system.time(deep_melt_reshape2 <- reshape2::melt(deep_list))
head(deep_melt_reshape2)
```

Although unlikely to encounter such large or deeply nested lists in practice, these artificial examples serve to illustrate that `reshape2::melt()` is not particularly efficient in unnesting large nested lists to data.frames. 

## Bind to wide data.frame

The option `how = "bind"` unnests a nested list to a *wide* data.frame and is used to unnest nested lists containing repeated entries of the same variables. To illustrate, we consider the `pokedex` dataset included in the `rrapply`-package, a nested list containing various property values for each of the 151 original Pokémon available (in .json) from https://github.com/Biuni/PokemonGO-Pokedex.

```{r}
## all 151 Pokemon
str(pokedex, list.len = 3)

## single Pokemon entry
str(pokedex[["pokemon"]][[1]])
```

Calling `rrapply()` with `how = "bind` expands each Pok&#233;mon sublist as a single row in a wide data.frame. The 151 rows are stacked and aligned by matching variable names, with missing entries replaced by `NA`'s (similar to `data.table::rbindlist(..., fill = TRUE)`). Note that any nested variables, such as `next_evolution` and `prev_evolution`, are unnested as wide as possible into individual data.frame columns similar to repeated application of `tidyr::unnest_wider()` to a data.frame with nested list-columns. 

```{r}
rrapply(pokedex, how = "bind")[, 1:9] |>
  head()
```

By default, the list layer containing the repeated observations is identified by the minimal depth detected across leaf elements. This option can be overridden by the `coldepth` parameter in the `options` argument, which can be useful to unnest nested sublists, such as `next_evolution` or `prev_evolution`. In addition, setting `namecols = TRUE` in the `options` argument includes the parent list names associated to each row in the wide data.frame as individual columns `L1`, `L2`, etc.

```{r}
## bind prev/next evolution columns
rrapply(
  pokedex, 
  how = "bind",
  options = list(coldepth = 5, namecols = TRUE)
) |>
  head(n = 10)
```

### Common alternatives

Several common alternatives used to unnest lists containing repeated entries include `data.table::rbindlist()`, `dplyr::bind_rows()`, and `tidyr`'s dedicated rectangling functions `unnest_longer()`, `unnest_wider()` and `hoist()`. 

The first two functions are primarily aimed at binding lists of data.frames or lists of lists, but are not meant for nested lists containing multiple levels of nesting, such as `pokedex`:

```{r, error = TRUE, message = FALSE}
library(dplyr)

## simple list of lists
lapply(pokedex[["pokemon"]], `[`, 1:4) |>
  bind_rows() |> 
  head()

## complex nested list (error)
bind_rows(pokedex[["pokemon"]])

## simple list of lists
lapply(pokedex[["pokemon"]], `[`, 1:4) |>
  data.table::rbindlist() |>
  head()

## complex nested list (error)
data.table::rbindlist(pokedex[["pokemon"]])
```

The rectangling functions in the `tidyr`-package offer a lot more flexibility. A similar data.frame as returned by `rrapply(pokedex, how = "bind")` can be obtained by repeated application of `tidyr::unnest_wider()`:

```{r}
library(tidyr)
library(tibble)

as_tibble(pokedex) |>
  unnest_wider(pokemon) |>
  unnest_wider(next_evolution, names_sep = ".") |>
  unnest_wider(prev_evolution, names_sep = ".") |>
  unnest_wider(next_evolution.1, names_sep = ".") |>
  unnest_wider(next_evolution.2, names_sep = ".") |>
  unnest_wider(next_evolution.3, names_sep = ".") |>
  unnest_wider(prev_evolution.1, names_sep = ".") |>
  unnest_wider(prev_evolution.2, names_sep = ".") |>
  head()
```

The option `how = "bind"` in `rrapply()` is less flexible as it always expands the nested list to a data.frame that is *as wide as possible*. On the other hand, the flexibility and interpretability in `tidyr`'s rectangling functions come at the cost of increased computational effort, which can become a bottleneck when unnesting large nested lists: 

```{r}
## large replicated pokedex list 
pokedex_large <- list(pokemon = do.call(c, replicate(1500, pokedex[["pokemon"]], simplify = FALSE)))

system.time({
  rrapply(pokedex_large, how = "bind")
})

## unnest first layers prev_evolution and next_evolution
system.time({
  as_tibble(pokedex_large) |>
    unnest_wider(pokemon) |>
    unnest_wider(next_evolution, names_sep = ".") |>
    unnest_wider(prev_evolution, names_sep = ".") 
})
```

**Remark**: in the chained calls to `unnest_wider()` above, we only unnest the first layer of the `next_evolution` and `prev_evolution` list-columns, and not any of the resulting children list-columns, which would only further increase computation time.

To extract and unnest sublists at deeper levels of nesting in the list, such as `next_evolution`, we manually set the `coldepth` parameter in the `options` argument, as also demonstrated above:

```{r}
system.time({
  ev1 <- rrapply(
    pokedex_large, 
    condition = \(x, .xparents) "next_evolution" %in% .xparents,
    how = "bind",
    options = list(namecols = TRUE, coldepth = 5)
  )
})
head(ev1)
```

The same unnested version of the `next_evolution` sublists can be obtained by mixing several calls to `unnest_wider()` and `unnest_longer()`:

```{r}
system.time({
  ev2 <- as_tibble(pokedex_large) |>
    unnest_wider(pokemon) |>
    unnest_longer(next_evolution) |>
    unnest_wider(next_evolution, names_sep = "_") |>
    select(id, next_evolution_num, next_evolution_name)
})
head(ev2)
```

In the context of the current example, a more efficient approach is to combine `unnest_wider()` with `hoist()`. The disadvantage is that we need to manually specify the exact locations of the elements that we wish to hoist from the nested list:

```{r}
system.time({
  ev3 <- as_tibble(pokedex_large) |>
    unnest_wider(pokemon) |>
    hoist(next_evolution, 
          name.1 = list(1, "name"),
          name.2 = list(2, "name"),
          name.3 = list(3, "name")
    ) |>
    select(id, name.1, name.2, name.3)
})
head(ev3)
```

Using `rrapply()`, the same result can be obtained by adding a call to `reshape()` (or alternatively e.g.  `tidyr::pivot_wider()` or `data.table::dcast()`) by converting from a long to a wide data.frame:

```{r}
system.time({
  ev4 <- rrapply(
    pokedex_large, 
    condition = \(x, .xparents) "next_evolution" %in% .xparents,
    how = "bind", 
    options = list(namecols = TRUE, coldepth = 5)
  ) 
  ev5 <- reshape(
    ev4[, c("L2", "L4", "name")],
    idvar = "L2",
    timevar = "L4",
    v.names = "name",
    direction = "wide"
  )
})
head(ev5)
```

### Additional examples

We conclude this section by replicating some of the data rectangling examples presented in the `tidyr` vignette: https://tidyr.tidyverse.org/articles/rectangle.html. The example nested lists are all conveniently included in the [repurrrsive](https://CRAN.R-project.org/package=repurrrsive)-package.

#### GitHub Users

```{r}
library(repurrrsive)

## nested data
str(gh_users, list.len = 3)

## unnested version
rrapply(gh_users, how = "bind") |>
  as_tibble()
```

#### GitHub repos

```{r}
## nested data
str(gh_repos, list.len = 2)

## unnested version
rrapply(gh_repos, how = "bind") |>
  as_tibble()
```

#### Game of Thrones characters

```{r}
## nested data
str(got_chars, list.len = 3)

## unnested version
rrapply(got_chars, how = "bind") |>
  as_tibble()
```

#### Sharla Gelfand’s discography

```{r}
## nested data (first element)
str(discog[1], list.len = 3)

## unnested version (excluding deeply nested sublists)
discs <- rrapply(
  discog,
  condition = \(x, .xpos) length(.xpos) < 5,
  f = \(x) ifelse(is.null(x), NA, x),  ## replace NULLs
  how = "bind"
)
as_tibble(discs)

## unnest labels sublists 
labels <- rrapply(
  discog,
  condition = \(x, .xparents) "labels" %in% .xparents,
  how = "bind",
  options = list(coldepth = 5, namecols = TRUE)
)
as_tibble(labels)

## merge disc id's with labels
merge(
  x = data.frame(L1 = rownames(discs), disc_id = discs[, "id"]),
  y = labels, 
  by = "L1", 
  sort = FALSE
) |>
  as_tibble()
```

# Data.frame to nested list

As a demonstrating example, we reconsider the long data.frame from the first section obtained after melting the renewable energy shares of all Western European countries:

```{r}
renewable_energy_melt_west_eu <- rrapply(
  renewable_energy_by_country,
  condition = \(x, .xparents) "Western Europe" %in% .xparents,
  how = "melt"
) 
head(renewable_energy_melt_west_eu, n = 10)
```

For certain tasks, it may be necessary to convert this data.frame back to a nested list object, e.g. to write the data to a JSON- or XML-object or for some tree visualization purpose. Writing a recursive function to reconstruct the nested list can prove to be quite time-consuming and error-prone. 

In this context, the `unlist()` function has an inverse counterpart `relist()` that reconstructs a nested list from the unlisted vector. The `relist()` function always requires a `skeleton` nested list to repopulate, which can make it difficult to use in practice, as such a `skeleton` object is for instance unavailable for the current example. In particular, the melted data.frame contains only a subset of the original list elements, so we can not use the original list as a template object without filtering nodes from the original list as well.

## Unmelt to nested list

To address this difficulty, `rrapply()` includes the dedicated option `how = "unmelt"` that performs the inverse operation of `how = "melt"`. No skeleton object is needed in this case, only a plain data.frame in the format returned by `how = "melt"`. To illustrate, we can convert the melted data.frame above to a nested list as follows:

```{r}
rrapply(
  renewable_energy_melt_west_eu, 
  how = "unmelt"
) |>
  str(give.attr = FALSE)
```

**Remark 1:** `how = "unmelt"` is based on a *greedy* approach parsing data.frame rows as list elements starting from the top of the data.frame. That is, `rrapply()` continues collecting children nodes as long as the parent node name remains unchanged. If, for instance, the goal is to create two separate nodes (on the same level) with the name `"Western Europe"`, these nodes should not be listed directly after one another in the melted data.frame as `rrapply()` will group all children under a single `"Western Europe"` list element.

**Remark 2:** Internally, `how = "unmelt"` reconstructs a nested list from the melted data.frame and subsequently follows the same conceptual framework as `how = "replace"`. Any other function arguments, such as `f` and `condition` can be used in exactly the same way as when applying `how = "replace"` to a nested list object.

**Remark 3:** `how = "unmelt"` does (currently) not restore the attributes of intermediate list nodes and is therefore not an exact inverse of `how = "melt"`. The other way around always produces the same result:

```{r}
renewable_energy_unmelt <- rrapply(renewable_energy_melt_west_eu, how = "unmelt")
renewable_energy_remelt <- rrapply(renewable_energy_unmelt, how = "melt")

identical(renewable_energy_melt_west_eu, renewable_energy_remelt)
```

In terms of computational effort, `rrapply()`'s `how = "unmelt"` can be equally or more efficient than `relist()` even though there is no template list object that can be repopulated. This is highlighted using the large list objects generated previously:

```{r}
## large deeply nested list (2^18 elements)
##  benchmark timing with rrapply
system.time(deep_unmelt <- rrapply(deep_melt, how = "unmelt")) 

## benchmark timing with relist
deep_unlist <- unlist(as.relistable(deep_list))
system.time(deep_relist <- relist(deep_unlist))
```

```{r}
## large shallow list (10^6 elements)
## benchmark timing with rrapply 
system.time(shallow_unmelt <- rrapply(shallow_melt, how = "unmelt")) 

## benchmark timing with relist
shallow_unlist <- unlist(as.relistable(shallow_list))
system.time(shallow_relist <- relist(shallow_unlist))
```

**Note**: the unmelted lists are not exactly identical to the original nested lists, since `how = "unmelt"` uses the placeholder names `1`, `2`, `3`, etc. in the melted data.frames to name the nodes in the newly constructed lists, whereas the name attributes in the original lists are all empty. By removing all names from the unmelted lists, they become identical to their original counterparts:

```{r}
## remove all list names
deep_unmelt_unnamed <- rrapply(
  deep_unmelt,
  f = unname,
  classes = "list",
  how = "recurse"
)
## check if identical
identical(unname(deep_unmelt_unnamed), deep_list)
```

```{r}
## remove all list names
shallow_unmelt_unnamed <- rrapply(
  shallow_unmelt,
  f = unname,
  classes = "list",
  how = "recurse"
)
## check if identical
identical(unname(shallow_unmelt_unnamed), shallow_list)
```

[^1]: The `renewable_energy_by_country` dataset is publicly available at the [United Nations Open SDG Data Hub](https://unstats-undesa.opendata.arcgis.com/datasets/)

[^2]: Note that `rrapply()` imposes a different column order than `reshape2::melt()` and the `"value"` column may follow slightly different coercion rules, but other than that the melted data.frames are the same.






